---
title: "MultiLevel"
author: "Erik Ketterer and Sagan Kakkar"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preliminaries
```{r libraries}
library(tidyverse)
library(gridExtra)
library(haven)
library(lme4)
```

# About data
The dataset we are going to use was created by the people over at the UCLA Institute for Digital Research and Education (IDRE). 

The dataset measures schools and teachers, as well as students. The researchers are interested in how the organizational structure of schools influences the performance of students, or how teacher characteristics such as experience, IQ, or teaching style have an impact on student learning.

Therefore, the level 1 observational unit is each student, and level 2 observational unit is school.

# Load data
```{r}
Multidata <- read_dta("https://stats.idre.ucla.edu/stat/examples/imm/imm23.dta")
```

# Research Question

Research Question 1: Does the time spent on homework each week affect a student's math score?

Research Question 2: Does the association between time spent doing homework and math score depend on the type of school (private or public)?

# Parameter of Interest and Fixed Effect/Random Effect

Therefore, the parameters we will use are 'homework' (time spent on homework each week) and 'public' (1=public school, 0 = others). And our response variable is 'math' (math score for each student).

Fixed effects are the time spent on their homework and the type of school.

Random effect is the school, indicated by 'schid' (school ID).

## Parameter fine-tuning
We are not going to do any parameter tuning here.
```{r}
head(mlmdata)
```

# Exploratory Data Analysis 

## Univariate Plots

```{r}
Multidata %>%
  summarize(overallmean = mean(math, na.rm = TRUE),
            overallsd = sd(math, na.rm = TRUE))

math_all <- ggplot(data = Multidata, mapping = aes(x = math)) +
  geom_histogram(binwidth = 3, fill = "navy", color = "white") +
  labs(x = "Math Score",
       y = "Number of Students") 

math_all
```

Based on the graph above, the distribution of math scores is approximately normal. The data is roughly symmetric about a score of 51.  We can see that the majority of scores are between 41 and 64.


```{r}
math_agg <- Multidata %>%
  group_by(schid) %>%
  summarize(mean = mean(math, na.rm = TRUE),
            sd = sd(math, na.rm = TRUE)) %>%
  ggplot(mapping = aes(x = mean)) +
     geom_histogram(binwidth = 2, fill = "navy", color = "white") +
     labs(x = "Mean Math Score",
          y = "Number of Schools") 

math_agg
```

Similar to the previous graph, the distribution of mean math scores per school is nearly normal; it is roughly symmetric around a score of 50. It seems that the majority of mean scores are between 45 and 57.

```{r}
# Plot with both univariate figures
grid.arrange(math_all,math_agg,ncol=1)
```


```{r}
# Changing indicator variable to text
Multidata <- Multidata%>%
    mutate(SchoolType = ifelse(public == 1, "Public", "Private"))

data_lev2 <-
  Multidata %>%
  group_by(schid) %>%
  filter(row_number() == 1)

# Creating proportions and table
table(data_lev2$SchoolType)
prop.table(table(data_lev2$SchoolType))

# Plot
inst <- ggplot(data = data_lev2, mapping = aes(x = SchoolType)) +
  geom_bar(fill = "navy", color = "white") +
  labs(x = "Type of School (Private vs Public)",
       y = "Frequency")
inst
```


The proportions and chart above show that private schools have a lesser frequency (0.35) than public schools (0.65) and because of this the results may not be fully representative of the population. 


```{r}
hwAll <- ggplot(data = Multidata, mapping = aes(x = homework)) +
  geom_histogram(binwidth = 1, fill = "navy", color = "white") +
  labs(x = "Time Spent on Homework (0-7)",
       y = "Number of Students") 

hwAll
```

The graph above shows a strong right skew. We note that most students spend an hour on homework each week. 

# Bivariate Plots

```{r}
bp <- ggplot(data = Multidata, mapping = aes(y = math)) +
  geom_boxplot(binwidth = 2, fill = "navy", color = "black") +
  labs(x = "Math Scores",
       y = "Number of Students") +
  facet_grid(. ~SchoolType)
bp
```

The spread of the private vs public math scores and the number of students is about the same for both types of schools, with the median for private schools (59) being noticeable higher than that for public (47) schools. 

```{r}
plot <- ggplot(data = Multidata, mapping = aes(y = math, x = homework))+
  geom_point()+
  geom_smooth(method = "lm", se = F)

plot
```

We can assume linearity between amount of time spent on homework and math scores based on the graph above.  There is a clear positive linear trend displayed by the blue line.

Lattice Plot:

```{r}
ggplot(Multidata,aes(x=homework,y=math,color = SchoolType)) +
  geom_point() + geom_smooth(method="lm",color="black") +
  facet_wrap(~schid,ncol=5) +  
  theme(strip.text.x=element_blank())  +
  labs(x="Time Spent on Homework (Hours)",y="Math Scores")
```

Similar evidence of linearity to the above plot shown in these individual plots per school.  These plots also show that private schools have higher math scores on average than do public.


## Model Building

### Unconditional means model; which we will call Model a (This is the model with no covariates and random intercepts only.) Model a is given below:

$$\text{Level 1: } Y_{ij} = a_i + \epsilon_{ij} \text{ where } \epsilon_{ij} \sim N(0, \sigma^2)$$

$$\text{Level 2: } a_i = \alpha_0 + u_i \text{ where } u_i \sim N(0, \sigma_u^2)$$

a. Write the composite model.

$$\text Y_{ij} = \alpha_0 + u_i + \epsilon_{ij} \text{ where } \epsilon_{ij} \sim N(0, \sigma^2) \text{, } u_i \sim N(0, \sigma_u^2)$$


```{r}
modela <- lmer(math ~ 1 + (1|schid), REML = F, data = Multidata)
summary(modela)
```

#### Intraclass Correlation Coefficient

```{r}
rho <- (24.85)/(24.85+81.24)
rho
```

About 23.4% of the total variability in math scores is due to differences between schools.  This value being above 10% (or rho = 0.10) gives us evidence to use a multilevel modeling technique.

### Model 1: Add homework as a level 1 covariate

$$\text{Level 1: } Y_{ij} = a_i + b_{i}x_{i,homework} + \epsilon_{ij} \text{ where } \epsilon_{ij} \sim N(0, \sigma^2)$$

$$\text{Level 2: } a_i = \alpha_0 + u_i $$
$$b_i = \beta_0 + v_i \text{ where } \begin{bmatrix} v_i \\ u_i \end{bmatrix} \sim N(\begin{bmatrix} 0 \\ 0 \end{bmatrix}, \begin{bmatrix} \sigma_u^2 \\ \rho\sigma_u\sigma_v & \sigma_v^2 \end{bmatrix})$$

a. Write the composite model.

$$\text Y_{ij} = \alpha_0 + \beta_0x_{i,homework} + u_i + v_ix_{i,homework}+ \epsilon_{ij} \text{ where } \epsilon_{ij} \sim N(0, \sigma^2) \text{, } \begin{bmatrix} v_i \\ u_i \end{bmatrix} \sim N(\begin{bmatrix} 0 \\ 0 \end{bmatrix}, \begin{bmatrix} \sigma_u^2 \\ \rho\sigma_u\sigma_v & \sigma_v^2 \end{bmatrix})$$

```{r}
model1 <- lmer(math ~ homework + (1+homework|schid), REML = F, data = Multidata)
summary(model1)
```

$$ \hat{\alpha_0} = 46.3 = \text{the mean math score before } $$

```{r}
model1.reduced <- lmer(math ~ homework + (1|schid), REML = F, data = Multidata)
summary(model1.reduced)
```



```{r}
model4 <- lmer(math ~ homework + public + (1+homework |schid), REML = F, data = Multidata)
summary(model4)
```



## Model Comparisons

WRITE HO AND H1

```{r}
# Drop-in-deviance
anova(model2, model4, test = "Chisq")
```

Test Statistic: 4.1955
P-value:        0.04053
Decision:   Fail to reject that the reduced model is sufficient
Conclusion: The model with hoemwork and school type (public) is preferred to the model with only homework.


# Model Analysis

Goodness of fit?

```{r}
summary(model4)
```


## Conclusion in Context

ADD


