---
title: "MultiLevel"
author: "Sagan Kakkar"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Preliminearies
```{r libraries}
library(tidyverse)
library(gridExtra)
library(haven)
library(lme4)
```

```{r readData}
Multidata <- read_dta("C:\\Users\\skakk\\Downloads\\imm23(2).dta")
```

# Exploratory Data Analysis 

## Univariate Plots

```{r}
Multidata %>%
  summarize(overallmean = mean(math, na.rm = TRUE),
            overallsd = sd(math, na.rm = TRUE))

math_all <- ggplot(data = Multidata, mapping = aes(x = math)) +
  geom_histogram(binwidth = 3, fill = "navy", color = "white") +
  labs(x = "Math Score",
       y = "Number of Students") 

math_all
```

Based on the graph above, the distribution of math scores is not normal, but we can use the central limit theorem to approximate normality due to large sample size. 


```{r}
math_agg <- Multidata %>%
  group_by(schid) %>%
  summarize(mean = mean(math, na.rm = TRUE),
            sd = sd(math, na.rm = TRUE)) %>%
  ggplot(mapping = aes(x = mean)) +
     geom_histogram(binwidth = 2, fill = "navy", color = "white") +
     labs(x = "Mean Math Score",
          y = "Number of Schools") 

math_agg
```
Similar to the previous graph, the distribution of mean math scores per school is not normal but we can use the central limit theroem to approximate normality due to large sample size. 



```{r}
# Plot with both univariate figures
grid.arrange(math_all,math_agg,ncol=1)
```


```{r}
# Changing indicator variable to text
Multidata <- Multidata%>%
    mutate(SchoolType = ifelse(public == 1, "Public", "Private"))

data_lev2 <-
  Multidata %>%
  group_by(schid) %>%
  filter(row_number() == 1)

# Creating proportions and table
table(data_lev2$SchoolType)
prop.table(table(data_lev2$SchoolType))

# Plot
inst <- ggplot(data = data_lev2, mapping = aes(x = SchoolType)) +
  geom_bar(fill = "navy", color = "white") +
  labs(x = "Type of School (Private vs Public)",
       y = "Frequency")
inst
```


The proportions and chart above show that private schools have a lesser frequency (0.35) than public schools (0.65) and because of this the results may not be fully representative of the population. 


```{r}
hwAll <- ggplot(data = Multidata, mapping = aes(x = homework)) +
  geom_histogram(binwidth = 1, fill = "navy", color = "white") +
  labs(x = "Time Spent on Homework (0-7)",
       y = "Number of Students") 

hwAll
```

The graph above shows a strong right skew, and we can also note that most students spend an hour on homework each week. 

# Bivariate Plots

```{r}
bp <- ggplot(data = Multidata, mapping = aes(y = math)) +
  geom_boxplot(binwidth = 2, fill = "navy", color = "black") +
  labs(x = "Math Scores",
       y = "Number of Students") +
  facet_grid(. ~SchoolType)
bp
```

The spread of the private vs public math scores and the number of students is about the same for both types of schools, with the median for private schools (59) being noticeable higher than that for public (47) schools. 

```{r}
plot <- ggplot(data = Multidata, mapping = aes(y = math, x = homework))+
  geom_point()+
  geom_smooth(method = "lm", se = F)

plot
```
COMMENT NEEDED

Though it looks like this may be better fit under a Poisson distribution, we will assume linearity in terms of practical model building.  We have not learned multilevel modeling for Poisson. 

Lattice Plot:

```{r}
ggplot(Multidata,aes(x=homework,y=math,color = SchoolType)) +
  geom_point() + geom_smooth(method="lm",color="black") +
  facet_wrap(~schid,ncol=5) +  
  theme(strip.text.x=element_blank())  +
  labs(x="Time Spent on Homework (0-7)",y="Math Scores")
```
COMMENT NEEDED

Similar evidence of linearity to the above plot shown in these individual plots per school.  These further show that private schools have higher math scores on average than do public.


## Model Building

```{r}
model1 <- lmer(math ~ 1 + (1|schid), REML = F, data = Multidata)
summary(model1)

model2 <- lmer(math ~ homework + (1+homework|schid), REML = F, data = Multidata)
summary(model2)

model2.reduced <- lmer(math ~ homework + (1|schid), REML = F, data = Multidata)
summary(model2.reduced)

model3 <- lmer(math ~ homework + ses + (1+homework+ses|schid), REML = F, data = Multidata)
summary(model3)

model4 <- lmer(math ~ homework + public + (1+homework|schid), REML = F, data = Multidata)
summary(model4)
```



## Model Comparisons

```{r}
# Drop-in-deviance
anova(model2, model4, test = "Chisq")
```

Test Statistic: 4.1955
P-value:        0.04053
Decision:   Fail to reject that the reduced model is sufficient
Conclusion: The model with hoemwork and school type (public) is preferred to the model with only homework.


# Model Analysis

```{r}
summary(model4)
```
The mean math score before accounting for the amount of time spent on homework and the type of school (private or public) is 49.065. 
The estimated variance within school deviations is 53.34.
The estimated variance between school deviations is 56.25.
The estimated variance between school deviations based on the amount of time spent on homework is 16.37.

# Intraclass Coefficient

```{r}
rho <- (56.25+16.37)/(53.34+56.25+16.37)
rho

rho_a <- (56.25)/(56.25+53.34+16.37)
rho_b <- (16.37)/(16.37+53.34+56.25)

rho_a
rho_b
```

44.6% of the variation in math scores is due to the differences between schools.
13.0% of the variation in math scores is due to differences between time spent on homework. 



